Syntax and Semantics of Indexed Abstract Binding Trees

### 2015-11-08 paper.tex
fix typo
taking a stab at an intro

\title{Abstract Binding Trees}
\date{}
\author{Jon Sterling and Darin Morrison}
\maketitle

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{\tau}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees (abts) is inductively defined in four rules:

@@

Each argument
%
\begin{tikzcd}[ampersand replacement = \&]
  {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
    \arrow[r, "\IMode{\gamma_i}"]
  \& {\IMode{%
    P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
  }}
\end{tikzcd}
%
is the exponential transpose of the composite,
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_u}
    }}
      \arrow[r, "\IMode{\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}}"]
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}}
      \arrow[r, "\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P}"]
    \& {\IMode{P_{\tau_i}}}
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}},\, \Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}}"]
    \& {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}},\, \Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}}"]
    \& {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}

>>>>

\title{Syntax and Semantics of Indexed Abstract Binding Trees}
\date{}
\author{Jon Sterling and Darin Morrison}
\maketitle

Abstract binding trees (abts) are a generalization of abstract syntax trees
where operators may express variable binding structure as part of their
arities.  Originally inspired by Martin-L\"of's theory of arities
\cite{martin-lof:1984}, unisorted abts have been deployed successfully as the
uniform syntactic framework for several implementations of Constructive Type
Theory, including Nuprl \cite{constable:1986}, MetaPRL \cite{hickey:2003} and
JonPRL \cite{jonprl:2015}.

In \emph{Practical Foundations for Programming Languages} \cite{harper:2016},
Robert Harper develops the multi-sorted version of abstract binding trees and
proposes an extension to include families of operators indexed by
\emph{symbols}, which are, unlike variables, subject to only
distinctness-preserving renaming and not substitution; furthermore, symbols do
not appear in the syntax of abts, and are only introduced as parameters to
operators. This generalization to \emph{indexed abts} is essential for a
correct treatment of programming languages with open sums (e.g. ML's
\texttt{exn} type), as well as assignable references.

In parallel, M.\ Fiore and his collaborators have developed the categorical
semantics for several variations of second-order algebraic theories
\cite{fiore:1999, fiore:2005, fiore-hur:2010, fiore-mamoud:2010}; of these, the
simply sorted second order algebraic theories treated by Fiore et al.\ are
equivalent to abstract binding trees augmented with a notion of second-order
variable (metavariable).

The contribution of this paper is the development of the syntax and semantics
of \emph{indexed multi-sorted abts}, an extension of second order universal
algebra to support symbol-indexed families of operators. Additionally, we have
developed the categorical semantics for indexed abts formally in Constructive
Type Theory using the Agda proof assistant.

@@

Let the judgment $\IsAbt{\Omega}{\Upsilon}{\Gamma}{M}{\tau}$ presuppose
$\IsMetaCtx{\Omega}$, $\IsSymCtx\Upsilon$, $\IsVarCtx\Gamma$ and $\IsSort{\tau}$,
meaning that $M$ is an abstract binding tree of sort $s$, with metavariables in
$\Omega$, parameters in $\Upsilon$, and variables in $\Gamma$. Let the judgment
$\IsBTm{\Omega}{\Upsilon}{\Gamma}{E}{v}$ presuppose $\IsValence{v}$.  Then, the
syntax of abstract binding trees is inductively defined in four rules:

@@

Each argument
%
\begin{tikzcd}[ampersand replacement = \&]
  {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P}}
    \arrow[r, "\IMode{\gamma_i}"]
  \& {\IMode{%
    P_{\tau_i}^{\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}}
  }}
\end{tikzcd}
%
is the exponential transpose of the composite,
\[
  \begin{tikzcd}[ampersand replacement = \&, sep=8em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\IMode{\Tuple{\pi_1\circ\pi_1,\phi_i,\psi_i}}"]
    \& {\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i}_P}}
      \arrow[r, "\IMode{\SemBrackets{\Omega\triangleright\Upsilon,\vec{\sigma}_i\parallel\Gamma,\vec{\tau}_i\vdash M_i:\tau_i}_P}"]
    \& {\IMode{P_{\tau_i}}}
  \end{tikzcd}
\]
where $\phi_i,\psi_i$ are defined as follows:
\begin{gather*}
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\phi_i\,}{\,\Tuple{\pi_u\circ\pi_1}_{u\in\Dom{\Upsilon}},\, \Tuple{\pi_u\circ\pi_1\circ\pi_2}_{u\in\Dom{\vec{\sigma}_i}}}"]
    \& {\IMode{S^{\{\Upsilon,\vec{\sigma}_i\}}}}
  \end{tikzcd}\\
  \begin{tikzcd}[ampersand replacement = \&, sep=14em]
    {\IMode{\SemBrackets{\Omega\triangleright\Upsilon\parallel\Gamma}_P
       \times\Yoneda{\vec{\sigma}_i\parallel\vec{\tau}_i}
    }}
      \arrow[r, "\Define{\psi_i\,}{\,\Tuple{\pi_x\circ\pi_1}_{x\in\Dom{\Gamma}},\, \Tuple{\pi_x\circ\pi_1\circ\pi_2}_{x\in\Dom{\vec{\tau}_i}}}"]
    \& {\IMode{V^{[\Gamma,\vec{\tau}_i]}}}
  \end{tikzcd}
\end{gather*}
